A Classical Density-Functional Theory for Describing Water Interfaces

### 2011-08-30 paper/project.tex
Changing units etc in figures, fixes to rods, text

Classical density-functional theory (DFT) provides an efficient way to evaluate the
free energy using the density of the fluid. The free energy is a functional of the 
density function and is independent of any external potentials\cite{ebner1976}. 
Mermin\cite{Mermin} gives a theorem extending that of Hohenberg and Kohn\cite{HK} 
to nonzero temperatures:
\begin{equation}
  A(T) = \underset{n(r)}{min}\left\{ F[n(r),T] + \int V_{ext}(r) n(r) d^3r\right\}
\end{equation}
This is the basis for classical DFT, which can be used with a variety of external potentials, 
from the simple interaction with a 
hydrophobic wall to the complex strong interaction with hydrophilic solutes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory and Methods}
We introduce a new classical density functional for water that
predicts the \needsworklater{vapor density, liquid density, bulk
  modulus, and bulk surface tension, as well as allows for
  liquid-vapor coexistence}.  The Helmholtz free energy functional is
based on SAFT, and is composed of the usual terms:
\begin{equation}
  F[n] = F_\textit{id}[n] + F_\textit{hs}[n] + F_\textit{assoc}[n] +
F_\textit{disp},
\end{equation}
where $F_\textit{id}$ is the ideal gas free energy, $F_\textit{hs}$ is
a hard-sphere free energy, $F_\textit{assoc}$ is the free energy of
association, and $F_\textit{disp}$ is the dispersion energy.  In
addition, a chemical potential term is used, so we actually work in
the grand canonical ensemble.  In the following sections, we will
introduce the terms of this functional.

\subsection{Ideal gas functional}
The first term is the ideal gas free energy functional,
which incorporates effects of kinetic energy.  The ideal gas
functional is given by
\begin{equation}\label{idealgas}
  F_{id}[n] = k_B T \int n(\xx)\left( \ln{n(\xx)} - 1\right) d\xx
\end{equation}
where n(\xx) is the density of water molecules.  The ideal gas free
energy functional has the property that the contact density at a hard
surface $n_c$ is proportional to the pressure on that surface,
according to the equation
\begin{equation}
  p = n_c k_BT \:,
\end{equation}
which also leads to the property that the free energy of solvation of
a hard solute is proportional to its volume in the limit of small
solutes.  These properties are retained by the total functional,
provided the remaining terms are \emph{purely
  nonlocal}~\cite{ashcroft?}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/saturated-liquid-density}
\end{center}
\caption{The theoretical versus experimental saturated liquid density
  versus temperature.  }
\label{fig:saturated-liquid-density}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/temperature-versus-density}
\end{center}
\caption{The theoretical versus experimental temperature versus density. }
\label{fig:saturated-liquid-density}
\end{figure}

@@

The calculated heat of vaporization at atmospheric pressure is
$\Delta H_{vap}$= 39.41 kJ/mol at the normal boiling point of water,
373 K. This compares very well to the experimental
value measured by NIST ($\Delta H_{vap}$= 40.65 kJ/mol\cite{nistwater}), with 
an error of only a few percent.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/surface-tension}
\end{center}
\caption{The theoretical versus experimental surface tension
  versus temperature.  }
\label{fig:surface-tension}
\end{figure}

@@

\fixme{I'm not sure if this paragraph belongs here.  Somewhere I think
  it's worth discussing any of these homogeneous plots we include, and
  this might be the right place.  We also might want to include one or
  more of them earlier in the text.  A general rule is that any figure
  that is included should be referenced in the text.  If there's
  nowhere in the text that it fits, then it probably doesn't belong in
  the paper....  So I'm not removing the follwoing bits since it's up
  to you, Jess, whether to remove the figures.  We should
  \emph{definitely} remove some of them.}  For reference, we plot the
predicted vapor pressure (see Figure~\ref{fig:equation-of-state}) and
saturated liquid density (see
Figure~\ref{fig:saturated-liquid-density}) against experimental
values.  To find these densities, we use the ``common tangent'' rule,
as illustrated in Figure~\ref{fig:homogeneous}.

\begin{table}
\begin{tabular}{ccc}
  Fitting Parameters& \hspace{2em} & Experimental Values\\
\begin{tabular}{cc}
  $R$ & 2.3 $a_o$ \\
  $A$ & -0.0100909 Hartree \\
  $\gamma$ & 5\\
  $V$ & 515.792 $a_o^3$  \\
  $b$ & 0.75 $a_o$ \\
  $\mu$ & -0.014063 Hartree 
\end{tabular}
&& 
\begin{tabular}{cc}
 $n_l$ & 4.939 $\times 10^{-3} a_o^{-3}$\\
 $n_v$ & 1.141 $\times 10^{-7} a_o^{-3}$\\
 $\sigma$ & ??? \\
 $B$ & $7.435\times 10^{-5}$ Hartree$/a_o^3$\\
 $T$ & 298.15 K \\
\end{tabular}
\end{tabular}

\caption{Empirical parameters of optimized functional and experimental
  values matched, where $a_o$ is the Bohr radius, and $\sigma$ is the
  surface tension.  This table is not yet complete, but it should
  be soon...  We'll also need to think about the best units to use
  here.  Atom units might be best, but it also might be better to
  consistently use eV and angstrom or something.  I'm not sure about
  putting $k_BT$ data in the table.  Maybe it belongs in the caption?}
\label{tab:parameters}
\end{table}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/energy-vs-diameter}
\end{center}
\caption{ Energy versus diameter for a single hydrophobic rod
immersed in water. This should have an asymptote equal to the surface
tension. (FIX - 0.4 nm?) }
\label{fig:energy-vs-diameter}
\end{figure}

@@

Figure \ref{fig:densityresolution} shows the density profiles for three different
values of the lattice spacing resolution. The density oscillations around a single
rod of radius 0.5 nm look nearly identical for all resolutions except for the first peak. 
This contact density is important to get correct as it \fixme{relates to energy}. The 
contact density calculated using the medium resolution is within less than 4\% of the value at
high resolution, while the low resolution contact density is off by about 5\%. There is some
uncertainty in the value of contact density due to the different lattice spacings. The total
free energy calculations compare much more favorably. The medium resolution is within 
0.5\% and the low resolution is within 1\% of the high resolution free energy.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/density-rods-in-water}
\end{center}
\caption{ Density profile for two hydrophobic rods (diameter 1.0 nm) at a 
separation of 0.8 nm. The scale on the right is density in g/mL.}
\label{fig:density-rods}
\end{figure}

>>>>

Classical density-functional theory (DFT) provides an efficient way to evaluate the
free energy using the density of the fluid. The free energy is a functional of the 
density function and is independent of any external potentials\cite{ebner1976}. 
Mermin\cite{mermin1965thermal} gives a theorem extending that of Hohenberg and 
Kohn\cite{hohenberg1964inhomogeneous} to nonzero temperatures:
\begin{equation}
  A(T) = \underset{n(r)}{min}\left\{ F[n(r),T] + \int V_{ext}(r) n(r) d^3r\right\}
\end{equation}
This is the basis for classical DFT, which can be used with a variety of external potentials, 
from the simple interaction with a 
hydrophobic wall to the complex strong interaction with hydrophilic solutes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory and Methods}
We introduce a new classical density functional for water that
predicts the \needsworklater{vapor density, liquid density, bulk
  modulus, and bulk surface tension, as well as allows for
  liquid-vapor coexistence}.  The Helmholtz free energy functional is
based on SAFT, and is composed of the usual terms:
\begin{equation}
  F[n] = F_\textit{id}[n] + F_\textit{hs}[n]  +
F_\textit{disp}[n]+ F_\textit{assoc}[n],
\end{equation}
where $F_\textit{id}$ is the ideal gas free energy, $F_\textit{hs}$ is
a hard-sphere free energy, $F_\textit{assoc}$ is the free energy of
association, and $F_\textit{disp}$ is the dispersion energy.  In
addition, a chemical potential term is used, so we actually work in
the grand canonical ensemble.  In the following sections, we will
introduce the terms of this functional.

\subsection{Ideal gas functional}
The first term is the ideal gas free energy functional,
which incorporates effects of kinetic energy.  The ideal gas
functional is given by
\begin{equation}\label{idealgas}
  F_{id}[n] = k_B T \int n(\xx)\left( \ln{n(\xx)} - 1\right) d\xx
\end{equation}
where n(\xx) is the density of water molecules.  The ideal gas free
energy functional has the property that the contact density at a hard
surface $n_c$ is proportional to the pressure on that surface,
according to the contact value theorem
\begin{equation}\label{contactvaluethm}
  p = n_c k_BT \:,
\end{equation}
which also leads to the property that the free energy of solvation of
a hard solute is proportional to its volume in the limit of small
solutes.  These properties are retained by the total functional,
provided the remaining terms are \emph{purely
  nonlocal}~\cite{ashcroft?}.

@@

% \begin{figure}
% \begin{center}
% \includegraphics[width=\columnwidth]{figs/saturated-liquid-density}
% \end{center}
% \caption{The theoretical versus experimental saturated liquid density
%   versus temperature.  }
% \label{fig:saturated-liquid-density}
% \end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/temperature-versus-density}
\end{center}
\caption{The theoretical versus experimental temperature versus density. }
\label{fig:temperature-vs-density}
\end{figure}

@@

The calculated heat of vaporization at atmospheric pressure is
$\Delta H_{vap}$= 39.41 kJ/mol at the normal boiling point of water,
373 K. This compares well to the experimental
value measured by NIST ($\Delta H_{vap}$= 40.65 kJ/mol\cite{nistwater}), with 
an error of only a few percent.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/surface-tension}
\end{center}
\caption{The theoretical versus experimental surface tension
  versus temperature. The experimental data is taken from NIST.\cite{nistwater} }
\label{fig:surface-tension}
\end{figure}

@@

\fixme{I'm not sure if this paragraph belongs here.  Somewhere I think
  it's worth discussing any of these homogeneous plots we include, and
  this might be the right place.  We also might want to include one or
  more of them earlier in the text.  A general rule is that any figure
  that is included should be referenced in the text.  If there's
  nowhere in the text that it fits, then it probably doesn't belong in
  the paper....  So I'm not removing the follwoing bits since it's up
  to you, Jess, whether to remove the figures.  We should
  \emph{definitely} remove some of them.}  For reference, we plot the
predicted vapor pressure (see Figure~\ref{fig:equation-of-state}) and
coexistence curve (see
Figure~\ref{fig:temperature-vs-density}) against experimental
values.  To find the densities, we use the ``common tangent'' rule,
as illustrated in Figure~\ref{fig:homogeneous}.

% \begin{table}
% \begin{tabular}{ccc}
%   Fitting Parameters& \hspace{2em} & Experimental Values\\
% \begin{tabular}{cc}
%   $R$ & 2.3 $a_o$ \\
%   $A$ & -0.0100909 Hartree \\
%   $\gamma$ & 5\\
%   $V$ & 515.792 $a_o^3$  \\
%   $b$ & 0.75 $a_o$ \\
%   $\mu$ & -0.014063 Hartree 
% \end{tabular}
% && 
% \begin{tabular}{cc}
%  $n_l$ & 4.939 $\times 10^{-3} a_o^{-3}$\\
%  $n_v$ & 1.141 $\times 10^{-7} a_o^{-3}$\\
%  $\sigma$ & ??? \\
%  $B$ & $7.435\times 10^{-5}$ Hartree$/a_o^3$\\
%  $T$ & 298.15 K \\
% \end{tabular}
% \end{tabular}
% 
% \caption{Empirical parameters of optimized functional and experimental
%   values matched, where $a_o$ is the Bohr radius, and $\sigma$ is the
%   surface tension.  This table is not yet complete, but it should
%   be soon...  We'll also need to think about the best units to use
%   here.  Atom units might be best, but it also might be better to
%   consistently use eV and angstrom or something.  I'm not sure about
%   putting $k_BT$ data in the table.  Maybe it belongs in the caption?}
% \label{tab:parameters}
% \end{table}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/energy-vs-diameter}
\end{center}
\caption{ Energy versus diameter for a single hydrophobic rod
immersed in water. This should have an asymptote equal to the surface
tension at room temperature, and it agrees well with the surface tension in Figure 
\ref{fig:surface-tension}. (FIX - 0.4 nm?) }
\label{fig:energy-vs-diameter}
\end{figure}

@@

Figure \ref{fig:densityresolution} shows the density profiles for three different
values of the lattice spacing resolution. The density oscillations around a single
rod of radius 0.5 nm look nearly identical for all resolutions except for the first peak. 
This contact density is important as it will be proportional to the pressure on the 
rod\ref{contactvaluethm}. The 
contact density calculated using the medium resolution is within less than 4\% of the value at
high resolution, while the low resolution contact density is off by about 5\%. There is some
uncertainty in the value of contact density due to the different lattice spacings. The total
free energy calculations compare much more favorably. The medium resolution is within 
0.5\% and the low resolution is within 1\% of the high resolution free energy.

@@

% \begin{figure}
% \begin{center}
% \includegraphics[width=\columnwidth]{figs/density-rods-in-water}
% \end{center}
% \caption{ Density profile for two hydrophobic rods (diameter 1.0 nm) at a 
% separation of 0.8 nm. The scale on the right is density in g/mL.}
% \label{fig:density-rods}
% \end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/rods-diagram}
\end{center}
\caption{ Cross section schematic for two hydrophobic rods immersed in water.
The radius and separation are shown.}
\label{fig:rods}
\end{figure}
